---
title: "Thyroid Cancer Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Thyroid Cancer Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/zaraavagyan/weathercsv/discussion\>*

### Reference:

*\<Avagyan, Z. (2017). Weather CSV [Data set]. Kaggle. https://www.kaggle.com/datasets/zaraavagyan/weathercsv\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r load dataset}
# Load dataset using read.csv
ThyroidCancerData <- read.csv("thyroid_dataset.csv", colClasses = c(
  Age = "numeric",
  Gender = "factor",
  Smoking = "factor",
  Hx_Smoking = "factor",
  Hx_Radiothreapy = "factor",
  Thyroid_Function = "factor",
  Physical_Examination = "factor",
  Adenopathy = "factor",
  Pathology = "factor",
  Focality = "factor",
  Risk = "factor",
  T = "factor",
  N = "factor",
  M = "factor",
  Stage = "factor",
  Response = "factor",
  Recurred = "factor"
))

# Preview the structure and a few rows
str(ThyroidCancerData)
head(ThyroidCancerData)
View(ThyroidCancerData)

```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Frequency tables for categorical variables
table(ThyroidCancerData$Gender)
table(ThyroidCancerData$Smoking)
table(ThyroidCancerData$Hx_Smoking)
table(ThyroidCancerData$Hx_Radiothreapy)
table(ThyroidCancerData$Thyroid_Function)
table(ThyroidCancerData$Physical_Examination)
table(ThyroidCancerData$Adenopathy)
table(ThyroidCancerData$Pathology)
table(ThyroidCancerData$Focality)
table(ThyroidCancerData$Risk)
table(ThyroidCancerData$Stage)
table(ThyroidCancerData$Response)
table(ThyroidCancerData$Recurred)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean, Median, Mode (custom function) for Age
mean(ThyroidCancerData$Age)
median(ThyroidCancerData$Age)

# Mode function
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
get_mode(ThyroidCancerData$Age)
```

## Measures of Distribution
```{r MOD}
# Measures of Central Tendency
# Standard Deviation, Variance, Range, and Summary for Age
sd(ThyroidCancerData$Age)          # Standard deviation
var(ThyroidCancerData$Age)         # Variance
range(ThyroidCancerData$Age)        # Range (min and max)
summary(ThyroidCancerData$Age)      # Min, 1st Qu., Median, Mean, 3rd Qu., Max

# Cross-tabulations (example relationships)
table(ThyroidCancerData$Gender, ThyroidCancerData$Stage)
table(ThyroidCancerData$Smoking, ThyroidCancerData$Stage)
table(ThyroidCancerData$Focality, ThyroidCancerData$Stage)
table(ThyroidCancerData$Risk, ThyroidCancerData$Recurred)
```

## ANOVA
```{r ANOVA}
# Chi-Square Test between Smoking and Stage
chisq.test(table(ThyroidCancerData$Smoking, ThyroidCancerData$Stage))

chisq.test(table(ThyroidCancerData$Gender, ThyroidCancerData$Recurred))

# ANOVA test for Age across different Stages
anova_result <- aov(Age ~ Stage, data = ThyroidCancerData)
summary(anova_result)
```

## Plots
```{r Plots}
# Load libraries
library(ggplot2)

# Histogram of Age
ggplot(ThyroidCancerData, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Age", x = "Age", y = "Count")

# Bar plot for Gender
ggplot(ThyroidCancerData, aes(x = Gender)) +
  geom_bar(fill = "lightgreen") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count")

# Bar plot for Stage
ggplot(ThyroidCancerData, aes(x = Stage)) +
  geom_bar(fill = "salmon") +
  labs(title = "Cancer Stage Distribution", x = "Stage", y = "Count")

# Boxplot of Age across Stage
ggplot(ThyroidCancerData, aes(x = Stage, y = Age, fill = Stage)) +
  geom_boxplot() +
  labs(title = "Age Distribution across Cancer Stages", x = "Stage", y = "Age") +
  theme(legend.position = "none")
```

# Preprocessing & Data Transformation
## Missing Values
```{r Missing values}
# Check if there are any missing values
any(is.na(ThyroidCancerData))

# Summary of missing values per column
colSums(is.na(ThyroidCancerData))

# Visualize missing values (optional)
# Install if needed: install.packages("Amelia")
library(Amelia)
missmap(ThyroidCancerData, main = "Missing Values Map", col = c("red", "green"), legend = FALSE)

```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load required package
library(caret)

# Set random seed for reproducibility
set.seed(123)

# Split data: 80% training, 20% testing
trainIndex <- createDataPartition(ThyroidCancerData$Stage, p = 0.8, list = FALSE)

trainData <- ThyroidCancerData[trainIndex, ]
testData <- ThyroidCancerData[-trainIndex, ]

# Check dimensions
dim(trainData)
dim(testData)
```

